Convergence rate of MAP estimates \\ for the exponential family

### 2021-02-23 main.tex
new stuff

\usepackage[round]{natbib}
\renewcommand{\bibname}{References}
\renewcommand{\bibsection}{\subsection*{\bibname}}

@@

\paragraph{Exponential Family.}
The exponential family member with sufficient statistic $T$ and natural parameter $\natp$ is the model
\begin{align}
    p(X|\natp) &= \exp( \natp^\top T(X) - \logpart(\natp)) \; ,
\end{align}
with log-partition function 
\begin{align}
    \logpart(\natp) = \log \int e^{\natp^\top T(x)} dx \; .
\end{align}
Recall that $\logpart$ verifies the two following identities 
\begin{align}
    \nabla\logpart(\natp) &=  \expect[p(X|\natp)]{T(X)} =: \meanp \\
    \nabla^2 \logpart(\natp) &= \Cov_\natp[T(X)] > 0
\end{align}
where $\meanp$ is called the mean parameter.
The second identity entails that $\logpart$ is strictly-convex.

@@

>>>>

\usepackage[round]{natbib}
\renewcommand{\bibname}{References}
\renewcommand{\bibsection}{\subsection*{\bibname}}ni

@@

\paragraph{Exponential Family.}
The exponential family member with sufficient statistic $T$ and natural parameter $\natp$ is the model
\begin{align}
    p(X|\natp) &= \exp( \natp^\top T(X) - \logpart(\natp)) \; ,
\end{align}
with log-partition function 
\begin{align}
    \logpart(\natp) = \log \int e^{\natp^\top T(x)} dx \; .
\end{align}
Recall that $\logpart$ verifies the two following identities.
\begin{align}
    \nabla\logpart(\natp) &=  \expect[p(X|\natp)]{T(X)} =: \meanp \\
    \nabla^2 \logpart(\natp) &= \Cov_\natp[T(X)] > 0
\end{align}
where $\meanp$ is called the mean parameter.
The second identity entails that $\logpart$ is strictly-convex.

@@

\subsection{A big calculus in this specific setting}

Indeed we know the distribution of $\mu$ it is linked with the chi-square distribution. We also know the expression of the excess of risk. So we could do a big calculus to get the exact expression of the averaged excess of risk.

Recall the expression of the Gamma integral
\begin{align}
  \Gamma(\alpha):=\int_0^\infty t^{\alpha-1} \exp(-t) d t
\end{align}
Recall also that
\begin{align}
    \int x^q \exp(-nx^p) dx
    &= n^{-(q+1)/p} \int (n^{1/p}x)^q \exp(-(n^{1/p}x)^p) d(n^{1/p}x)\\
    &= n^{-(q+1)/p} \int u^q \exp(-u^p) du\\
    &= n^{-(q+1)/p} \int t^{q/p} \exp(-t) d(t^{1/p})\\
    &= n^{-(q+1)/p} \int p^{-1} t^{(q+1)/p - 1} \exp(-t) dt\\
    &= n^{-(q+1)/p} p^{-1} \Gamma((q+1)/p).
\end{align}
Finally if there is problem in zero with the logarithm appearing in Eq. (75, at time of writing), we can have a Taylor expansion of the log 
\[
  \log(x) = \sum_{i=1}^\infty \frac{(-1)^{i+1}(x-1)^i}{i},
\]
and show convergence with some Fubini's theorem on series-integral.

\subsection{Generic approach}

In this subsection, I detailed a generic approach to get non-asymptotic rates.
It is based on the calibration inequality Eq. (18, at time of writing).
The next step would be to detailed those derivations and have precise results for the specific case of only looking at the variance.

\paragraph{Calibration inequality.}
One of the main difficulty of the proof is that $\theta_n = (\nabla A)^{-1}(\mu_n)$ can be really big, meaning that we can not apply directly ``calibration inequality'' between the Bregman divergence ${\cal B}_{A^*}(\mu^*; \mu)$ and $\norm{\mu - \mu^*}^2$.
However, if we restrict ourselves to the domain $\theta'\in {\cal B}(0, R)$ with $R = \max(\norm{\theta}, \norm{\theta^*})$, we have that $A$ is $2^{-1}R^{-2}$-strongly convex. Therefore using Eq.~(18, at the time of writing), we have 
\begin{align}
    {\cal B}_{A^*}(\mu^*; \mu) 
    &\leq 2^{-1}R^{-2} \norm{\mu^* - \mu}^2\\
    &= 2 \max(\norm{\theta}, \norm{\theta^*})^2\norm{\mu^* - \mu}^2\\
    &= 2 \max(\norm{(\nabla A)^{-1}(\mu)}, \norm{(\nabla A)^{-1}(\mu^*)})^2\norm{\mu^* - \mu}^2.
\end{align}
With will proceed with this equation, we might get better equations. For example by considering a constant of strong convexity on the segment $[\norm{\theta}, \norm{\theta^*}]$ rather than the ball ${\cal B}(0, R)$.
The behavior of the max, will be similar to the behavior of the sum of the two elements since $\max(a, b) \leq a + b \leq 2\max(a, b)$, therefore, we proceed
with
\begin{align}
    {\cal B}_{A^*}(\mu^*; \mu) 
    &\leq 2 \norm{(\nabla A)^{-1}(\mu^*)}^2 \norm{\mu^* - \mu}^2
    + 2 \norm{(\nabla A)^{-1}(\mu)}^2 \norm{\mu^* - \mu}^2.
\end{align}
The first term is easy to control in expectation, see Eqs. (19-20, at time of writing), it is $\Var{T(X)} / n$. It is really the second part that we want to study. 
Because of the structure $\norm{(\nabla A)^{-1}(\mu)} \simeq \norm{\mu}^{-1}$
We have to take care of the event ``$\mu = n^{-1} \sum_{i\leq n} T(X_i)$ too small''.
We know that $\mu$ concentrate towards $\mu^*$ so this event should not happen too often.
So we want to study $\norm{(\nabla A)^{-1}(\mu)}^2 \norm{\mu - \mu^*}^2$.

\paragraph{Tail bound or average excess of risk?}
If we think with concentration inequality, we will get tail bound on the excess of risk, and not an average excess of risk.
However, recall that
\begin{align}
  \E[X] = \E[\int_{0}^{\infty} 1_{X > t} dt]
  = \int_{0}^\infty \E[1_{X>t}] dt
  = \int_0^\infty \mathbb{P}(X > t) dt.
\end{align}
So we can translate tail bound into average.

\paragraph{Concentration of a product.}
We want to study $\norm{(\nabla A)^{-1}(\mu)}^2 \norm{\mu - \mu^*}^2$, it is a product of two term, we can proceed with the fact that if $A$ and $B$ are two random variables from a sample space $(\Omega, \mu)$ to $\mathbb{R}$, we have
\[
    \{ \omega \in \Omega \vert A(\omega) B(\omega) > t\}
    \subset 
    \{ \omega \in \Omega \vert A(\omega) > s\}
    \cup
    \{ \omega \in \Omega \vert A(\omega) > t/s\}
\]
Which results from the fact that if $ab > t$ than either $a > s$, either $b > t/s$.
In our case, it means that for any $t > 0$, we have the deviation bound
\begin{align}
  \mathbb{P}(\norm{(\nabla A)^{-1}(\mu)}^2 \norm{\mu - \mu^*}^2 > t)
  &\leq \inf_{s>0} \mathbb{P}(\norm{(\nabla A)^{-1}(\mu)}^2 > 1/s)
  + \mathbb{P}(\norm{\mu - \mu^*}^2 > st)\\
  &= \inf_{s>0} \mathbb{P}(\norm{(\nabla A)^{-1}(\mu)} > s^{1/2})
  + \mathbb{P}(\norm{\mu - \mu^*} > (st)^{1/2}).
\end{align}
So we are left with the study of 
\begin{align}
  \mathbb{P}(\norm{\mu - \mu^*} > t)
  \qquad\text{and}\qquad 
  \mathbb{P}(\norm{(\nabla A)^{-1}(\mu)} > t).
\end{align}

\paragraph{Concentration on $\mu - \mu^*$.}
Most likely $T(X)$ is sub-Gaussian (do hand-derivations or look at works like \url{https://arxiv.org/pdf/1011.3027.pdf} or \url{https://arxiv.org/pdf/1902.03736.pdf}) and Bernstein concentration inequality leads to a tail bound of the type, with $\sigma^2$ a variance parameter 
($\sigma^2 = \Var_{\theta^*}(T(X))$)
\begin{align}
    \mathbb{P}_{(X_i)_{i\leq n}}(\norm{\mu - \mu^*} > t) \leq \exp(-n t^2 / \sigma^2).
\end{align}
Normally, in Bernstein inequality there is an additional term in $n$, but the convergence rates will not depend on this additional term, so I am omitting it for simplicity.
Note that when we will integrate that we will get
\begin{align}
    \int_0^\infty \mathbb{P}_{(X_i)_{i\leq n}}(\norm{\mu - \mu^*} > t^{-1/2}) dt
    &=\int_0^\infty \exp(-n t / \sigma^2) dt\\
    &= \sigma^2 n^{-1}\int_0^\infty \exp(- nt / \sigma^2) d(nt\sigma^{-2})\\
    &= \sigma^2 n^{-1} \int_0^\infty \exp(-x) dx =\sigma^2 n^{-1}.
\end{align}

\paragraph{Probability that $\mu$ is close to zero.}
Intuitively, we should study $\mathbb{P}(\norm{(\nabla A)^{-1}(\mu)} > t)$, which should relates to $\mathbb{P}(\norm{\mu} < t)$, which can be express exactly as we know the cumulative distribution function of the chi-square and that $\mu$ is a variant of the chi-square.
It seems to me that here we will not get optimal rates, because the calibration inequality is too harsh when $\mu$ goes to $0$, as I develop in the next paragraph.

\paragraph{Big picture.}
Intuitively, we have summarize in the part in $\mathbb{P}(\norm{(\nabla A)^{-1}(\mu)} < t)$, the problem that arise with ${\cal B}_{A^*}(\mu^*; \mu)$ when $\mu$ is small. 
We could make this intuition more formal with
\begin{align}
   {\cal B}_{A^*}(\mu^*; \mu)
   &= 1_{\norm{\mu} \geq s} {\cal B}_{A^*}(\mu^*; \mu)
   + 1_{\norm{\mu} < s} {\cal B}_{A^*}(\mu^*; \mu)\\
   &\leq \sup_{\norm{\mu'} \geq s}(\norm{(\nabla A)^{-1}(\mu')})^2 \norm{\mu - \mu^*}^2
   + 1_{\norm{\mu} < s} {\cal B}_{A^*}(\mu^*; \mu).
\end{align}
Suppose the $\sup$ is finite, the first term is easy to control.
The second term is what we should take care of. We have
\begin{align}
    1_{\norm{\mu} < s} {\cal B}_{A^*}(\mu^*; \mu) 
    &=  1_{\norm{\mu} < s} (A^*(\mu^*) - A^*(\mu) - <\nabla A^*(\mu), \mu^* - \mu>)\\
    &\leq  1_{\norm{\mu} < s} (A^*(\mu^*) - \inf_{\norm{\mu'} < s} A^*(\mu') + \norm{\nabla A^*(\mu)} \sup_{\norm{\mu''} < s} \norm{\mu^* - \mu''}).
\end{align}
We know that, under some strict convexity condition, $\nabla A^* = (\nabla A)^{-1}$ (similar to Eq. (25) at time of writing). Therefore we really have to study, with harsh notation
\begin{align}
    \int_{0}^s \norm{(\nabla A)^{-1}(\mu)} d\mathbb{P}(\norm{\mu} < t)
    =\int_{0}^s \norm{(\nabla A)^{-1}(\mu)} \mathbb{P}(\norm{\mu} \in [t, t+dt]).
\end{align}
It seems to me that with the precedent development we were implicitly studying
\begin{align}
    \int_{0}^s \norm{(\nabla A)^{-1}(\mu)}^2 d\mathbb{P}(\norm{\mu} < t),
\end{align}
which of course is worse.
The study can be done precisely as we know the distribution of $\mu$.
In terms of rates, in my current view, this path is tight if and only if $<\nabla A^*(\mu), \mu^* - \mu>$ do behave like $\norm{(\nabla A)^{-1}(\mu)}$ towards $0$. In terms of constants, we have done pretty harsh majoration in order to get back to $c\norm{\mu-\mu^*}$ with constants $c$ that are clearly suboptimal.

\paragraph{Tail bound or average risk?}
Note that, if $\mu$ is sub-Gaussian concentrating towards $\mu^*$,
we can give a simpler result than average risk, 
which is that we can that that 
with proba $1 - \mathbb{P}(\norm{\mu} < 1/2\norm{\mu^*}) = 1 - \exp(-c\norm{\mu^*}^2 n)$, 
we have a tail bound $\mathbb{P}(\norm{\mu}^{-2}\norm{\mu - \mu^*}^2 > t) = \exp(-cnt / \norm{\mu^*})$.
This means that we can say that 
with proba $1 - \exp(-c\norm{\mu^*}^2 n) - \delta$ for any $\delta >0$, 
we have that $B_{A^*}(\mu^*; \mu) < cn^{-1}\log(1/\delta) / \norm{\mu^*}$.
